# House prices as internal migration determinants.

```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
library(dygraphs)
library(mapview)
library(leafem)
library(tidyverse)
library(sf)
library(leaflet)
library(gganimate)
library(gifski)
library(png)
library(ggrepel)
library(scales)
library(plotly)
library(ggthemes)
library(xts)
library(zoo)
library(htmltools)
library(stringr)
library(broom)
library(ggeasy)
library(usmap) 

library(TSstudio)

```


From an economic perspective, house prices represent a main component of inflation, overall price level trends are currently a priority for economic policy in general and particularly relevant when studying residential mobility.

Empirical evidence reported by current academic research  [@https://doi.org/10.1002/psp.2412] as well as market experts [@s&p/case-shillerEconomicData2023], indicate that, beyond labor market incentives, housing affordability represent a key factor when considering the selection of a place to reside [@FUSS201635].

In fact, real state markets may play a critical role for important social processes such as migration, both internal (within the same country)^[For the purposes of this investigation, internal migration is defined as the change in residence location from one particular Metropolitan area to another] or international. 

In this chapter, exploratory data analysis is conducted on house property values, focusing on metropolitan areas as the geographic scale across the US.  

Metropolitan areas are geographic units that are particularly suited to understand regional real state dynamics and provide a base scenario to study internal migration patterns within the US territory as well as specific regions such as the US-Mexico border.

## Data and methods.

The study strategy follows a spatial and temporal design, it is argued that analyzing housing costs evolution in a geographic context may reveal patterns that help to better understand residential mobility and deliver valuable insights about internal migration flows and the underlying regional economic incentives.

The primary spatial unit of analysis for this study is the Metropolitan Statistical area (MSAs). MSAs are geographical regions defined by the Office of Management and Budget (OMB) that consist of one or more counties that contain a core city of 50,000 or more inhabitants, and surrounding areas with significant social and economic ties [@ofr1990federalregister]. 

It is argued that Metropolitan statistical areas provide an appropriate level of detail, improving over the aggregate perspective that states offer or the traditional county boundaries scale, which lack the economic synergies and the broad social component that MSAs achieve by considering the interaction between a core and a set of adjacent communities. For this theoretical considerations MSAs were selected over other alternative geographical scales for this particular analysis. 

The study covers all 48 continental adjacent US states, excluding: District of Columbia, Alaska and non continental territories Hawaii and Puerto Rico.

A primary goal is to explore real state market fluctuations over time, to achieve this purpose we measure three concepts: house price levels, price appreciation and purchasing parity based on housing prices. 

To measure house price levels the main input variable is the quarterly seasonally adjusted house price index for the period 1995:Q1 (base year) to 2022:Q4 for 387 MSAs for which  Federal housing finance agency  (FHFA) price indices are available. 

The FHFA House Price Index (FHFA HPI) is a broad economic measure of the movement of single family house prices in the United States [@fhfaHousePriceIndex2022].  The data set analyzed in this chapter contains purchase only  transactions, the aim is to capture the price level fluctuations a feature estimated by using sales price data as opposed to appreciation. (which considers a repet transaction methodology.

To measure the underlying rate of appreciation of a given house price, the Freddie Mac House price index (FMHPI), is incorporated. This is a repeat transaction index, which compares the price of the same property over two or more transactions and is explicitly designed to capture appreciation while holding constant basic factors such as physical location, climate, housing type, etc.,between transactions. [@freddiemacFreddieMacHouse2023].

By construction, therefore, the repeat transaction requirement exclude new homes.

Whereas HPI in data sets such as FHFA, measures current price level not considering transactions over the same property.  

To complement the perspective obtained by the Federal housing finance agency HPI data and the Freddie Mac HPI,  the housing component of the regional price parities (RPPs) indicator for year 2022 (the latest release as the time of writing this investigation) is included.

This is an official indicator reported by the Bureau of economic analysis (BEA) which measure the differences in price levels across states and metropolitan areas for a given year and are expressed as a percentage of the overall national price level[@bureauofeconomicanalysisRegionalPriceParities2023]. 

This indicator is particularly informative of the population purchasing power, specially it allows price comparisons relative to the national average.

To capture demand side factors, Official population estimates from the US Census Bureau were obtained for years 2010-2020 [@uscensusbureauPopulationEstimatesEstimated2023, this is a variable that captures demographic dynamics overtime and serve as a proxy for housing demand.  

Per capita personal Income (PPI) by metropolitan area from the U.S. Bureau of Economic Analysis is also included.
[@bureauofeconomicanalysisPersonalIncomeCounty2023]. This variable contributes to capture inequality and financial barriers to access the housing market. In addition the income to house price ratio is useful to inform on the contextual factors affecting the decisions to move. 

A second goal is to emphasize the impact of monetary policy shifts on housing demand by examining time series data for interest rates. Specifically, we analyze how changes in interest rates influence the demand for home loans, which, in turn, has a significant effect on home prices.

Interest rates serve as a crucial economic indicator that directly affects the demand for housing. By exploring the dynamics of interest rates over time, we aim to shed light on the influential role they play in shaping housing market trends in the presence of external economic shocks, particularly important during the study period we document: a)  the effect of the so called "Great Recession" related to the sub-prime mortgage market collapse in 2009 and b) the global health emergency declaration of early 2020.

A third goal is to analyze non-conventional economic data^[Conventional economic data is defined for the purposes of this investigation as repositories generated by the Census Bureau, Bureau of economic analysis and federal agencies and institutions that provide estimates, projections and indicators related to the housing market], to understand the population sentiment towards the housing market, we propose the use of the Google trends tool. 

This tool provides insights into the relative popularity of search terms over a particular period of time and across different regions. And it has become a relevant source for academic research recently from areas in the social sciences field as wellas other disciplines such as natural sciences and public health. [@askitasTrendSpottingHousingMarket2016], [@avramescuNowcastingRomanianMigration2021], [@vazquezGRANDESDATOSGOOGLE2020], [@kamMonitoringDroughtAwareness2019].

By analyzing the search volume data obtained from Google, Google Trends allows us to observe and compare the changes in interest and search patterns for various topics using a score in the close domain [0-100], where 100 indicates the highest volume of searches for a particular and 0 indicates the lowest volume for the term at a precise moment in time and location.

By analyzing the search behavior of internet users, Google Trends offers a valuable resource for understanding the relative level of interest and attention that a topic or keyword generates within the online population. We analyze three important words that characterize equal number of aspects for the housing markets: affordable home, Foreclosure and eviction. 
This keywords allow to answer questions on where an when does people (Google users) are looking for this words, specially the term Foreclosure, serve as a proxy for home owners interest whereas eviction represents the renter´s segment. 


## Exploratory data analysis.

The analysis is organized in two stages: the first provides a national approach to the housing prices fluctuations for state and MSA scales. While the second focuses on the US Mexico border region.

When we analyze real estate market fluctuations in the context of affordability housing, understanding regional inflation patterns is crucial, in particular in this section the analysis starts by exploring the housing component for the regional price parity (RPP), this is an indicator developed by the Bureau of Economic Analysis (BEA) to measure price fluctuations at a regional scale. 

Examining this aggregate economic indicator it is possible to distinguish which states of the country exhibit the most expensive housing markets and therefore which areas of the country represent particular challenges either for households entering the market or those families already paying a mortgage. 


Regional price parties capture variations in the cost of living across different region of the country, this indicator is important because it helps to understand the relative purchasing power and living standards in different geographic areas and particularly when considering the housing component, it provides a measure to compare the real estate market conditions relative to the country.  

The RPP base level is the national average which takes a value of 100, so RPP values above this national threshold indicate relative more expensive states in terms of housing prices.

Figure 1 shows regional price parities by state relative the average US for 2021.  

**Figure 1 Regional price parities (RPP) by state 2022**
```{r, echo=FALSE, warning=FALSE,message=FALSE}

### Plot  For Regional price parities for the housing component

## Loading the data set.   REGIONAL PRICE PARITIES
# Source: 

prices<-read.csv("prices.csv")%>%
  mutate(type=as.factor(type), type=ifelse(type==0, "Below avg", "Above avg"))%>%
  filter(!row_number() %in% 9)%>%
  filter(state!="Alaska" &  state!="Hawaii")

### Plot  

rppp<-ggplot(prices, aes(x = reorder(state,rppindex), y=rppindex, fill = type))+ 
  geom_col()+  
  coord_flip()+
  theme_minimal()+
labs(fill="RPP")+
theme(legend.justification=c(1,0), legend.position=c(.9,.4))+
ylab("Regional Price Parities (RPP) 2022. Housing component. (%) Deviation")+
xlab("State")+
annotate("text", x="Alabama", y=50, size=3.1,label="US Housing RPP=100 per cent")+
  annotate("text", x="California", y=70, size=3.1,label="64.9")+
  annotate("text", x="Arizona", y=6, size=3.1,label="2.4")+   annotate("text", x="Texas", y=-8, size=3.1,label="-4.4")+ 
   annotate("text", x="NewMexico", y=-30, size=3.1,label="-27")+
scale_fill_manual(values=c("deepskyblue4","firebrick"), labels=c("Below average","Above average" ), name="")+
  theme(legend.position = "bottom", legend.title = element_blank()) 

ggplotly(rppp)%>%
  layout(legend = list(orientation = "h", x = 0.3, y = -0.2))

 #ggsave("rpp2022_housing_USstates.tiff", width = 25, height = 17, units = "cm",dpi=300)
```
Source: José Luis Manzanares analysis of Regional price parities data 2021 [@bureauofeconomicanalysisRegionalPriceParities2023]


It can be seen that the Us west coast border state of California has the highest housing price parity among the 48 states (excluding Hawaii and Alaska) for 2021. This  pattern suggest financial pressures for residents in west coast states that may explain the current out of state internal migration documented by academic research elsewhere [@blumenbergJobsHousingBalance2021] for states such as California.

In order to get a higher level of detail on the behavior of house price fluctuations,  metropolitan statistical areas around the country are now analyzed, considering the House price index developed by the Federal Home Loan Mortgage Corporation^[Freddie Mac, indicated as FM for further reference in this work], this measure is explicitly designed to capture appreciation of the housing market as discussed in the methodology section of this chapter.  

The data analysis strategy design aims to extract the appreciation over a two decade time frame for each metropolitan area, so, we are able to measure not only which metropolitan areas have registered the bigger price gains between 2000 and 2020 but also it´s annual rate of growth. 

Measuring appreciation using the HPI developed by a government sponsored enterprise such as the FM, is important for two reasons in the context of this investigation: I) It informs on the potential changes on home equity fluctuations over time with a geographic perspective, which expose real estate investment opportunities  II) It reflects existing financial barriers and/or opportunities to entry into the housing market for households looking to relocate, so is an appropriate economic indicator that helps to explain  current internal migration patterns.  

Figure 2 below shows the House Price index (HPI) distribution across the 387 U.S. Metropolitan statistical areas as of 2020. Each circle represents a HPI measure with its size corresponding to the HPI value.

The index depicts housing market prices taking as base year 1995. So higher values indicate a relative appreciation and therefore more expensive locations than those with smaller circles.   


**Figure 2 Metropolitan Statiscal areas House Price Index distribution 2020**
```{r,  warning=FALSE, message=FALSE, echo=FALSE}

## This datas set contains 387   MSA as of  2020 with geo atribututes and avg hpi  2020  base year is 1995   which means that HPI for uS is 100% in 1995. The avg_q  variable takes into account a mean for q1,q2,q3,q4  in the given year.

df<-readRDS("datageo2020.rds")%>%rename(Year=yr, MSA=msa.x, Annual.HPI.growth00.20=gwr00_20, State=state)

# Adding a factor  variable for border states 
df <- st_as_sf(df,  crs = 4326)
ca<-filter(df,State==" CA"|State==" CA (MSAD)")
# Interactive Map 

mapviewOptions(vector.palette=colorRampPalette(c("blue","orange","red")))

mapa<-mapview(df, map.types =  "OpenStreetMap.DE",cex="hpi20", zcol="hpi20", layer.name= "MSA House Price Index",legend = FALSE,homebutton =FALSE)
  
leafem::addMouseCoordinates(mapa)%>%setView(zoom = 3.6, -96.5716694, 39.1836082)
```
Source: Jose Luis Manzanares Rivera based on  House price index data from [@fhfaHousePriceIndex2022]



Empirical findings reveal a pronounced surge in housing prices, as gauged by the House Price Index (HPI), within geographically delineated metropolitan areas across the United States. Noteworthy regions exhibiting such trends encompass metropolitan statistical areas (MSAs) situated along the West Coast, the southeastern part of Florida, and the northwestern region. Over the span of the past three decades, these identified regions have witnessed substantial increase in housing values.

Among the states located on the West Coast, the State of California is particularly outstanding with areas of concern such as the Bay area where the so called silicon valley region represents a mayor labor market hub particularly for highly skilled workers associated with the technology sector. This geographic area holds substantial economic importance due to its concentration of technological innovation, fostering an environment  that attracts and sustains a skilled workforce with above average income which also translates into relevant tax revenue sources for the entire state. 

Furthermore within California, and right at the US southern border, the San Diego MSA, ranks among the top 10 most  expensive metropolitan areas of the entire state with a relative house price increase of 363.6 per cent in the last 25 years.

Using the above estimated pattern it si convenient determine **which metropolitan areas registered the highest price increments since base year 1995** according to the Federal House Finance HPI? 

Figure 3 illustrates the ranking of the nation's 20 foremost metropolitan areas based on their respective positions in the House Price Index (HPI) for the year 2020. It is important to note that these areas signify the highest relative increases in home prices compared to the base year of 1995, rather than indicating the absolute highest-priced urban population centers. 

**Figure 3 Top 20 Metropolitan areas by HPI increase.**
```{r, echo=FALSE}
#df<-readRDS("datageo2020.rds")

msa2022<-readRDS("msa2022.rds")%>%
  mutate(State=as.factor(ifelse(state==" CA"|state==" CA (MSAD)","California","Other state")), top2022=ifelse(hpi>=479.435,1,0))%>%
mutate(MSA=fct_reorder(msa, hpi))%>%
  rename(place_id=cbsa_code)


## The plot.

top22_hpi<-ggplot(msa2022, aes(hpi, MSA, fill=State))+
  geom_col()+
  scale_fill_manual(values = alpha(c("dodgerblue4", "gray")), name="")+
  labs(
    x = 'House Price index 2020. (1995=100 percent)',
    y = 'Metropolitan statiscal areas (MSAs)',
    alpha = 'Class',
    title = '')+ theme(legend.position="bottom")

ggplotly(top22_hpi)%>%
  layout(legend = list(orientation = "h", x = 0.3, y = -0.2))

```
Source: Jose Luis Manzanares Rivera based on  House price index data from [@fhfaHousePriceIndex2022]

**Figure 3a House price index annual growth rate 1995-2022**

```{r, include=FALSE}
#floflo<-data.frame(f=c(" FL"))

#rate<-readRDS("rate.rds")%>%mutate(geo=as.factor(ifelse(state%in%border_st$border,"USMXBorder", ifelse(state%in%floflo$f,"Florida","No Border"))))
                                         
#rate$msa[rate$msa=="Fort Lauderdale-Pompano Beach-Sunrise"] <- "Fort Lauderdale"

#rate$msa[rate$msa=="West Palm Beach-Boca Raton-Boynton Beach"] <- "West Palm Beach-Boca Raton"

#saveRDS(rate, "rate.rds")
```

```{r, echo=FALSE}
rate<-readRDS("rate.rds")
cols <- c( "#E74C3C","#154360", "#1E8449")
          
tsplot<-ggplot(rate, aes( Year, AnnualGrowth,color=geo, label = state))+
  geom_line() + 
facet_wrap(~msa, scales = "free_y")+
  geom_hline(yintercept=0)+
  geom_vline(xintercept=2009, linetype="dotdash")+
  scale_color_manual(values = cols, name="")+
  labs(y="Annual HPI growth rate (%)")
  
ggplotly(tsplot)%>%
  layout(legend = list(orientation = "h", x = 0.2, y = -0.1))
```
Source: Jose Luis Manzanares Rivera based on  House price index data from [@fhfaHousePriceIndex2022]

I would like to take on four patterns found when analyzing the top 20 metropolitan areas with the highest appreciation rates as measured by the annual growth on the HPI during the last 27 years  
I) The importance of the USMX border region. $40\%$ of the top 20 (MSAs) with the highest appreciation rates are located  at border states.

II) Within the US Mexico border, the impartance of California state is clear representing $75\%$ of the border MSA with highest HPI growth. In fact San Diego-Chula Vista-Carlsbad exhibits the highest annual increase in 2022 with a $18.55\%$  

II) The real state market in Florida shows the fastest growing level of prices with sharp increases particularly during the Trump administration, with Metropolitan areas such as West Palm Beach, North Port-Sarasota-Bradenton, Naples Marco Island and Tampa featurin double digit increases with amex growth of  32.77 the highest HPI annual growth in the nation.   

III) Traditional non high HPI areas in border states such as Texas are now starting to gain dynamism, the clear example in this reagion is the Austin-Roundrock-Georgetown Metropolitan area, where external economic factors such as the recent establishment of companies to support the electric vehicle industry have contribute to create speculative pressures on the real estate local markets.^[
In 2020, Tesla announced its decision to build a Gigafactory in Austin, which is expected to manufacture electric vehicles and employ thousands of workers. Suppliers, service providers, and other companies in the electric vehicle industry have begun to establish operations near the Gigafactory, creating additional job opportunities and further contributing to housing demand. The presence of a major company like Tesla often attracts real estate investors looking to capitalize on the economic growth and potential appreciation in property values. Increased investor interest can lead to higher demand for housing, further impacting the housing market dynamics] 



Although the above spatial patterns offer detailed empirical evidence to understand the incentives for population movements between metropolitan areas within states, one question that arises in the context of economic determinants of house prices is:  What role does available economic policy instruments have played to   shape the US real state markets in the last two decades of the current century? 

Certainly this period has witnessed at least two of the most remarkable and complex scenarios in economic recent history that deserve to be explained given its relevance for US society in general and households paying a mortgage in particular: I) the Great Recession, the economic downturn that occurred between December 2007 and June 2009. II) the COVID19 health emergency.

In accordance with conventional economic theory, a fundamental factor influencing the overall fluctuations in the aggregate price level is the cost of borrowing, commonly known as the interest rate. Interest rates are a core instrument of monetary policy that play a pivotal role in shaping economic dynamics and has significant implications for various sectors of the economy, including real estate markets. 

Therefore, understanding the evolution in time of interest rates is crucial for comprehending not only the aggregate price level fluctuations, but also its impacts on particular components, among which housing prices bear a specially important role households. 

We now turn to the analysis of this important macroeconomic determinant and begin by presenting the trayectory overtime for the Federal Funds effective rate. 
This rate serves as a benchmark or reference point for other short-term interest rates, exerting a considerable influence on the overall cost of borrowing and lending in the economy.

The importance of the Federal Funds effective rate (FFER) is particularly pronounced in relation to the housing sector and real estate markets. Firstly, it directly affects the cost of mortgage borrowing for homebuyers. Changes in the Federal Funds rate can lead to adjustments in mortgage interest rates, impacting affordability and demand for housing. Lower rates tend to stimulate borrowing and housing activity, while higher rates may dampen demand. This inverse relation is paramount to explain demand side factors that have been taking place in the US housing market as we will soon discover. 

Figure 4, presents the evolution of the FFER in the period from  to July 2000 to May 2023. 

**Figure 4. Monetary policy: The federal Funds Effective reate evolution July 2000 - May 2023**
```{r, echo = FALSE}

# FFER plot 

ts_FFER<-readRDS("ts_FFER.rds")

dygraph(ts_FFER, main = "Monetary policy: FFER")%>% 
dyRangeSelector()%>%
dyHighlight(highlightCircleSize = 3, 
highlightSeriesBackgroundAlpha = 0.2
)%>%
dySeries("FEDFUNDS", color = "#D35400",fillGraph = TRUE,label ="Federal Funds Effective Rate")%>%
dyAxis("y", label = "Percent")%>%
dyLegend(width = 300)%>%
dyEvent("2008-07-01", "QE Subprime mortgage market collapse", labelLoc = "top", color = "black")%>%
dyEvent("2020-02-01", "NBER Economic recession: QE -Ultra Low interest rates-", labelLoc = "top", color = "blue")%>%
  dyEvent("2022-02-01", "Monetary policy change ending quantitative easing", labelLoc = "top", color = "red")%>%
dyOptions(axisLineWidth = 1.5, fillGraph = TRUE, drawGrid = FALSE)

# Source: https://fred.stlouisfed.org/series/FEDFUNDS
```

Source: José Luis Manzanares Rivera based on data from Federal Reserve. [@boardofgovernorsofthefederalreservesystemFederalFundsEffective2023]

The representation above covers 23 years of Federal Funds effective rate fluctuations. Over the span of this period, monetary policy has experienced important changes with implications for the housing markets. 

The first, marked here with a benchmark date in reference to the sub-prime mortgage market collapse ^[Sub-prime mortgage is defined here as higher-risk loans granted to borrowers with lower creditworthiness], An episode in the recent economic history of the United States that embodies one of the primary determinants shaping the so called  "Great Recession."

Prior to this period, the Federal Reserve´s monetary policy had set the FFER at $5.26\%$ which its the highest interest levek that US society has experienced between July 2007 and the moment of the writing of this work (May 2023).

Following the sub-prime mortgage market collapse, which set the stage for the beginning of the global financial crisis, the US central bank, decided to implement measures via the its key monetary policy instruments, to stimulate the economy. 

Considering the substantial scale and breadth of the economic contraction, it was anticipated that the path to recovery would entail a protracted ascent. Consequently, the Federal Funds Effective Rate (FFER) underwent a deliberate reduction, plummeting to historically low levels, reaching a floor of $0.07\%$ by January 2014.

How does the housing market reacted to this unprecedented period of ultra low interest rates? 

Before moving on to answer that question, let´s study a second relevant shift in monetary policy. According to empirical evidence shown above,  the monetary policy stimulus period to promote economic recovery related to the Great Recession, lasted 7 years, (Jan 2009-Dec 2015).

However when quantitative easing^[For the purposes of the present investigation we use the term Quantitative easing to refer to a monetary policy tool employed by central banks to stimulate the economy and promote lending and investment. QE involves the central bank purchasing government bonds or other financial assets from commercial banks and other financial institutions. By doing so, the central bank injects new money into the financial system, increasing the money supply and lowering interest rates. The aim of quantitative easing is to encourage borrowing and spending, thereby boosting economic activity] seemed to be reversed; on January 31, 2020, Department of Health and Human Services (HHS) under the Public Health Service Act, declared a public COVID-19 health emergency. This declaration allowed the HHS to activate emergency response mechanisms, mobilize resources, and coordinate efforts to contain and mitigate the spread of the public heath threat within the country. 

Among those measures, as we now Know, there were significant mobility restrictions that infringed a toll on economic activity with emphasis in the labor market. Triggering a relevant economic slowdown, that required once more, a decisive intervention by the central bank authorities to prevent a deeper economic crisis and to stimulate a recovery.

In this unprecedented complex global economic setting, the Federal Reserve of the United States decided to move the FFER from a $2.4%$ in July 2019 to an ultra low  $0.05\%$ in April 2020, sending a clear signal to the markets  to stimulate spending and investment in a afford to correct the sharp decline in consumer spending associated with the labor market disruptions caused by the overall restrictive measures put in place during the global health emergency. 

This monetary policy was extended for almost two years and started to be normalized by april 2022. 


We may now turn to study the effects of this macroeconomic determinants and external shocks to the economy into the housing markets and its relation to cirrent internal migration patterns with a focus in specific regions of the country, such as the US-Mexico border states and metropolitan areas. 

Figure 5, below examines time series from the Federal Reserve financial accounts data to show the evolution of mortgage debt outstanding in relation to the housing stock market value, an important feature that offers a unique perspective to measure home equity growth a key indicator that reflects supply and demand dynamics for the housing market as a result of fluctuations registered on primary macroeconomic environment instruments such as the federal funds effective rates. 

Although this approach provides an nation wide perspective it is necessary to capture a base trend whose implications then can be examined using a detailed geographic scale. 


**Figure 5 Real estate at market value and rising equity 1980-2023**
```{r, echo=FALSE}
# Plots.
## Rising equity 
## loading data 

ts_b101<-readRDS("ts_b101.rds")
b101_e<-readRDS("b101_e.rds")

## Dygraph plots

dy_graph_equity <- list(
dygraph(ts_b101,  group= "a", main = "Real estate at market value")%>% 
dyRangeSelector()%>%
dyHighlight(highlightCircleSize = 3, 
highlightSeriesBackgroundAlpha = 0.2,
hideOnMouseOut = FALSE)%>%
dySeries("mortgage", label ="Mortgage Debt Outstanding", fillGraph = TRUE, color = "red")%>%
dySeries("value", label ="Real estate at market value", fillGraph = TRUE, color = "green")%>%
dyAxis("y", label = "(Trillion  USD, not)")%>%
dyLegend(width = 300)%>%
dyOptions(drawGrid = FALSE)%>% 
dyEvent("2008-07-01", "Subprime mortgage market collpse", labelLoc = "bottom", color = "purple")%>%
dyEvent("2020-04-01", "Quantitative Easing-Ultra Low interes rates-", labelLoc = "top", color = "blue"), 
  
dygraph(b101_e, group= "a", main = "Raising Equity")%>% dyRangeSelector()%>%
dyHighlight(highlightCircleSize = 3, 
highlightSeriesBackgroundAlpha = 0.2,
hideOnMouseOut = FALSE)%>%
dySeries("equity",fillGraph = TRUE, color = "#34495E", label = "Home Equity" )%>%
dyAxis("y", label = "House stock Equity (Trillion)")%>%
dySeries("Percent", label = "%" , color="#641E16",axis = 'y2', strokePattern = "dashed")%>%
dyAxis("y2", label = "Share of total housing stock value ")%>%
dyLegend(width = 300)%>%
dyOptions(drawGrid = FALSE))

htmltools::browsable(htmltools::tagList(dy_graph_equity))
```
Source: José Luis Manzanares Rivera based on data from [@federalreseveFinancialAccountsUnited2023]


For existing home owners, home equity has more than double during the last decade (from 2006Q4 to 2022Q4 with a 213% growth).

Why there was such growth in home equity in the nation during the last decade? 

As we have seen, monetary policy is partially responsible for this trend, given a sustained period of quantitative easing. (starting from December 2008 with a Federal reserve effective rate (FFFER) = $0.16\%$ down from $5.26\%$ in July 2007 -prior to sub prime mortgage market collapse- Followed by a second ultra low FFER period that was implemented to cope with a deceleration of the economy in the presence of the global public health emergency triggered  by the COVID-19 restrictions, starting at 2020Q1 with FFER = $0.05\%$ down from FFER=$2.40 \%$ in june 2019. 

Who benefited from this historically monetary policy scenario  characterized by ultra low interest rate? 

Specially people who buy homes in attracting real state markets areas with dynamic internal migration patterns, who manage to profit from this  historic ultra low interest rate period, by entering the market and securing a Fixed-rate mortgage just after the sub prime mortgage collapse, that is between December 2008 and November 2015.  

This movers took advantage of the economic stimulus provided the macroeconomic environment created with a monetary policy oriented for quantitative easing, a strategy  designed to lower the price of money in an effort to boost consumption and investment,among other things. For example, home equity for existing housing stock in the US in 2012Q1 was $10.25$T and it grew to $33.78$T in 10 years,  a 3X order of magnitude increase!

Lets now take a closer look froma geographic perspective of this real estate markets impacts. Figure 6 below 




```{r, echo=FALSE}

#top20_metros_inv<-readRDS("top20metros_invest.rds")%>%  mutate(USMXBorder=as.factor(ifelse(state%in%border_st$border,"USMXBorder","No Border")))%>% slice_max(t, n=20)

#saveRDS(top20_metros_inv, "top20_metros_inv.rds")

top20_metros_inv<-readRDS("top20_metros_inv.rds")%>%
  rename(Investment.Loans=t, MetropolitanArea=GEO_Name)%>%
  mutate(MSA=fct_reorder(MetropolitanArea, Investment.Loans))

inv_plot<-ggplot(top20_metros_inv, aes(Investment.Loans, MSA, fill = USMXBorder)) +
  geom_col() +
  scale_fill_manual(values = c("#AEB6BF", "#C0392B"), name="") +
  labs(title = "Top 20 Metropolitan Areas. Investment Home loans 2022-Q1",
       x = 'Total investment loans',
       y = 'Metropolitan Statistical area')

ggplotly(inv_plot)%>%
  layout(legend = list(orientation = "h", x = 0.3, y = -0.2))
```

Source: José Luis Manzanares River based on data from Freddie Mac Single-Family Loan-Level 2022Q1 [@freddiemachomeSingleFamilyLoanLevel2023]


